% ═══════════════════════════════════════════════════════════════════════════
%  EXPERIMENT I — Baseline single-run comparison
%  Paste Table~\ref{tab:comparison} immediately after this block.
%  Requires: \usepackage{booktabs, siunitx, amsmath, amssymb}
% ═══════════════════════════════════════════════════════════════════════════


% ───────────────────────────────────────────────────────────────────────────
%  Paragraph that introduces the table  (paste before \input{table})
% ───────────────────────────────────────────────────────────────────────────

Table~\ref{tab:comparison} summarises the steady-state tracking
performance of DQ-MPCC and the quaternion-based MPCC baseline over a
single \SI{100}{m} lap of the Lissajous reference path.
Both controllers share identical cost weights, constraints, prediction
horizon ($T_p = \SI{0.5}{s}$, $N = 50$ nodes), and control frequency
(\SI{100}{Hz}), so any difference in the metrics is attributable
exclusively to the state-space representation:
quaternion $+$ $\mathrm{SO}(3)$ error for the baseline,
versus dual quaternion $+$ $\mathrm{SE}(3)$ logarithm for DQ-MPCC.
Statistics are computed sample-wise over all $N_{\mathrm{ctrl}}$
control steps of the lap; the lap time $t_{\mathrm{lap}}$ is the primary
scalar indicator because $v_\theta$ is an optimised control variable
and a fixed-time criterion would artificially penalise the faster
controller.


% ───────────────────────────────────────────────────────────────────────────
%  <<<  INSERT \input{comparison_table} HERE  >>>
% ───────────────────────────────────────────────────────────────────────────


% ───────────────────────────────────────────────────────────────────────────
%  Paragraph that analyses the table  (paste after \input{table})
% ───────────────────────────────────────────────────────────────────────────

The results show a consistent advantage for DQ-MPCC across the
path-following metrics.
The mean position error is reduced by $17\%$
($\SI{0.64}{m}$ vs.\ $\SI{0.77}{m}$) and the peak drops from
\SI{2.27}{m} to \SI{1.65}{m}, confirming that the coupled
rotation-translation representation of $\mathrm{SE}(3)$ produces tighter
spatial tracking.
The contouring and lag errors follow the same trend
($-17\%$ and $-18\%$ respectively), indicating that DQ-MPCC stays closer
to the path both laterally and longitudinally.
The only metric where the baseline is marginally better is orientation
error ($\SI{0.52}{rad}$ vs.\ $\SI{0.74}{rad}$): the $\mathrm{SE}(3)$
logarithm couples rotation and translation in a single cost term, so
the optimiser allocates part of the orientation budget to reduce
positional error, accepting a slightly larger rotational residual.
The net effect on task performance is nonetheless positive:
DQ-MPCC achieves a mean progress speed of $\SI{5.81}{m/s}$
(vs.\ $\SI{4.95}{m/s}$) and completes the \SI{100}{m} lap in
$t_{\mathrm{lap}} = \SI{17.19}{s}$, a $\mathbf{16\%}$ reduction over
the baseline ($\SI{20.43}{s}$).
Despite the richer representation, the mean solver time increases only
from \SI{3.16}{ms} to \SI{5.06}{ms}---both well within the
\SI{10}{ms} budget of the \SI{100}{Hz} loop---confirming that the
$\mathrm{SE}(3)$ formulation is real-time feasible on the target hardware.
